Evolution of Processes in Delivering Reliable Software within the European Scientific Arena

### 2018-08-31 paper-ieee-trans.tex
added a possible answer to editor s comments for the lack of quantification

\begin{table*}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force            &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling             &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling              &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies                &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection             &       &       &       &$\surd$&       &$\surd$\\
Build/Testing Management Procedure &$\surd$&$\surd$&$\surd$&$\surd$&       &$\surd$\\
Software Product Metrics           &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition        &       &       &$\surd$&$\surd$&$\surd$&$\surd$\\
Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
Auto--generated Documentation      &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD) &       &       &       &       &       &$\surd$\\
\hline
\hline
\end{tabular}
\end{table*}

@@

The E--Infrastructure for Testing, Integration and Configuration of Software~\cite{etics}
({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
at addressing the challenges in producing quality software in distributed,
collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The framework
integrated different technologies and tools in order to provide automated configuration,
build and testing capabilities, as well as auto--generated documentation and
software metrics gathering -- such as Source Lines Of Code (SLOC), complexity and
number of defects/bugs \cite{etics} --. The {\sl ETICS} framework was the first automated
service for delivering quality software products in distributed environments like
the Grids.

@@

The European Middleware Initiative ({\sl EMI}, May 2010 -- Apr 2013)
\cite{cordis:emi} project joined the 4 major Grid middleware providers in
Europe at the time -- {\sl gLite}, {\sl UNICORE}, {\sl ARC} and {\sl dCache} --
with the goal to maintain and evolve the middleware focusing on extending their
interoperability and improving the reliability of the services. The
ISO/IEC 9126 \cite{iso-9126} standard was used in order to identify a set of
characteristics that needed to be present in the {\sl EMI} software products and
processes to be able to meet the {\sl EMI} quality requirements
\cite{emi-quality-model}.

For each software characteristic, a set of associated
metrics and Key Performance Indicators (KPIs) were identified and defined in
detail in the {\sl EMI} Metrics Specification \cite{emi-quality-model}. The project
leveraged the {\sl ETICS} service for the development, continuous integration and release management,
as well as for metric tracking, making queries on the collected data to display
them through a chart generation framework.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of various operational tools -- such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb} -- and the
management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
the production infrastructures. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers, such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software had to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defined a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The released software is tested in production environments through the
Staged--Rollout process, before its inclusion in the EGI's Cloud Middleware Distribution (CMD).
Selected resource providers are requested to install
the most updated stable versions of the software, and to give access to them to their users. The
Staged--Rollout process is key to detect and mitigate issues that could only
appear in production environments.

>>>>

\begin{table*}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force            &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling             &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling              &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies                &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection             &       &       &       &$\surd$&       &$\surd$\\
Build/Testing Management Procedure &$\surd$&$\surd$&$\surd$&$\surd$&       &$\surd$\\
Software Product Metrics           &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition        &$\surd$&       &$\surd$&$\surd$&$\surd$&$\surd$\\
Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
Auto--generated Documentation      &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD) &       &       &       &       &       &$\surd$\\
\hline
\hline
\end{tabular}
\end{table*}

@@

The E--Infrastructure for Testing, Integration and Configuration of Software~\cite{etics}
({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
at addressing the challenges in producing quality software in distributed,
collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The ETICS framework
integrated different technologies and tools in order to provide automated configuration,
build and testing capabilities, as well as auto--generated documentation and
software metrics gathering -- such as Source Lines Of Code (SLOC), complexity and
number of defects/bugs \cite{etics} --. The {\sl ETICS} framework was the first automated
service for delivering quality software products in distributed environments like
the Grids.

@@

The European Middleware Initiative ({\sl EMI}, May 2010 -- Apr 2013)
\cite{cordis:emi} project joined the 4 major Grid middleware providers in
Europe at the time -- {\sl gLite}, {\sl UNICORE}, {\sl ARC} and {\sl dCache} --
with the goal of maintaining and evolving the different Grid middleware solutions, 
focusing on extending their interoperability and improving the reliability of the services. The
ISO/IEC 9126 \cite{iso-9126} standard was used in order to identify a set of
characteristics that needed to be present in the {\sl EMI} software products and
processes to be able to meet the {\sl EMI} quality requirements
\cite{emi-quality-model}.

For each software characteristic, a set of associated
metrics and Key Performance Indicators (KPIs) were identified and defined in
detail in the {\sl EMI} Metrics Specification \cite{emi-quality-model}. The project
leveraged the {\sl ETICS} service for the development, continuous integration and release management,
as well as for metric tracking, making queries on the collected data in order to display
them through a chart generation framework.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of various operational tools -- such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb} -- and the
management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
the production infrastructures. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers, such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software had to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defined a set of requirements in different areas: documentation, deployment, security, information
model and operations.

The impact of the quality assurance activities was observed in the software provided by EMI to the EGI--InSPIRE project. Data reported in Table \ref{tab:emi} (see page 13 in \cite{emi-quality-model}) summarizes the EMI software quality evolution per project quarters (PQ) as evaluated by the EGI project with the help of the UMD QC. It is possible to observe the improvement of EMI releases over time that met the required quality criteria by considering the products that have passed the UMD QC, the products that have passed the EGI Stage Rollout (SR) phase (i.e. it consisted of testing the certified updates of the middleware by a subset of sites before made them available to all sites), and the products rejected because they did not meet the UMD QC.
\begin{table}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{EMI software quality evolution}
\label{tab:emi}
\centering
\begin{tabular}{lp{1cm}p{1cm}p{1cm}p{1cm}}
\hline
\hline
PQ & Release Products (RP) & N. RP Passed UMD QC & N. RP Passed SR & N. RP Failed UMD QC\\
\hline
\hline
5 & 30 & 27 & 27 & 0\\
6 & 30 & 28 & 26 & 2 \\
7 & 27 & 26& 24& 2\\
8 & 18 & 18 & 18 & 0\\
\hline
\hline
\end{tabular}
\end{table}

{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The released software is tested in production environments through the
SR process, before its inclusion in the EGI's Cloud Middleware Distribution (CMD).
Selected resource providers are requested to install
the most updated stable versions of the software, and to give access to them to their users. The
Staged--Rollout process is key to detect and mitigate issues that could only
appear in production environments.
